---
title: "fixedpoints: Basics"
author: "Simon Barthelmé"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r init,echo=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE, cache=FALSE, 
               comment=NA, verbose=TRUE, fig.width=4, fig.height=4,dev='png')
```

# Thinning a set of points

A set of points randomly drawn from a uniform distribution will typically clump a bit:

```{r}
library(spatstat)
pp <- rpoint(200) #200 random points
plot(pp)
```

Some of these points are really close together. For some applications it's preferable to have points that are at least a certain distance apart. We can get there by thinning the point set, removing iteratively points that are too close to their nearest neighbour:

```{r}

rm.point <- function(pp,eps=.05)
    {
        d <- nndist(pp) #Find distance to nearest neighbour
        if (any(d < eps)) #Any of them too close? 
            {
                chuck <- which.min(d) #Find the worst offender
                pp[-chuck] #Get rid of it
            }
        else
            {
                pp
            }
    }
```

Turning our iteration into a fixed point iteration we get a thinning algorithm:

```{r}
thin <- fp(rm.point)
plot(thin(pp))
plot(thin(pp,eps=.03))
```

# Optimisation by gradient descent

In gradient descent the minimum of a function $f(x)$ is found by iterating:

$$ x_{t+1} = x_t - \alpha \nabla f(x_t) $$

Here is an implementation using fp: 

```{r}

cost.fun <- function(x) .5*sum(x^2)
grad.cost <- function(x) x

gd.iter <- function(x,alpha=.2) x-alpha*grad.cost(x)
gd <- fp(gd.iter)+viewer(function(x) sprintf("Cost: %.3e",cost.fun(x)) %>% print,10,TRUE)
x0 <- rnorm(10)
gd(x0)

```

We can choose to stop the iteration once the cost is low enough:

```{r}
gd <- fp(gd.iter)+convergence(function(x,xp) cost.fun(xp),tol=1)

gd(x0)
#Change max. cost
gd <- gd+convergence(tol=.1)
gd(x0)
```

Or we could choose to stop the iteration once the cost has stopped decreasing:

```{r}
gd <- fp(gd.iter)+convergence(function(x,xp) abs(cost.fun(xp)-cost.fun(x)),tol=.01)
gd(x0)

```
# Region labelling

Region labelling is a  common task in image segmentation. The input is a black and white image (ie. the pixels have value 0/1), and the goal is to "label" each continuous white region, i.e. change the pixel value in this region from 1 to $i$, where i is the region's index.

It turns out to be very easy to implement region labelling as a fixed point iteration. Our iteration is to find the first pixel with value 1 in the image, and use a "bucket fill" to change the value of the surrounding white region. We iterate until there are no more such pixels.

Here is the iteration:

```{r}
library(imager)
region.label <- function(im)
    {
        #Get coordinates to all the pixels with value 1
        whitePix <- as.data.frame(im) %>% subset(value == 1)
        if (nrow(whitePix) > 0)
            {
                coord <- whitePix[1,c("x","y")] #first pixel we find 
                bucketfill(im,coord$x,coord$y,color=max(im)+1)
            }
        else
            {
                im
            }
    }

```

And here's the full iteration on an example:

```{r }
set.seed(1)
blobs <- imnoise(500,500) %>% isoblur(25) %>% threshold(0)
plot(blobs)
g <- fp(region.label)
plot(g(blobs))

```

The algorithm is easy to understand if we visualise the process:

```{r fig.width=10, fig.height=6}

show.iter <- function(im,iter)
    {
        regs <- as.data.frame(im) %>% dplyr::filter(value > 1)
        centers <- dplyr::group_by(regs,value) %>% dplyr::summarise(mx=mean(x),my=mean(y),label=unique(value)-1)
        plot(im,main=paste("Iter",iter))
        with(centers,text(mx,my,label,col="red"))
    }

layout(matrix(1:6,2,3,byrow=TRUE))
g <- fp(region.label)+viewer(show.iter,every=1)
g(blobs)
```
